Lazy Linearity for a Core Functional Language

### 2025-10-16 popl26.tex
up to recursive let

% Indeed, usage environments allow us to encode mutual exclusivity between
% alternative ways of consuming linear resources (between $\D$-vars and direct
% resource usage). 
Let-bound variables are the canonical example of a
$\Delta$-bound variable.
% , that is, let-bound variables bind expressions in which
% the resources required to type them are consumed lazily rather than eagerly.
%
Annotating let-bound variables with \ROUNDTWO{their} usage environment $\D$ delays the
consumption of resources to when the variables themselves are used and
guarantees the desired mutual exclusion property between the binder and the resources:
\[
\TypeLet
\]
The rule for (non-recursive) let bindings splits the linear environment in $\D$
and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
surprisingly, the resources $\D$ used to type $e$ are still available in the
environment to type the let body $e'$, alongside the unrestricted $x$ binding
annotated with the usage environment $\D$. In essence, the resources of $e$ being
available in $e'$ reflects how typing the body of a lazy binder
doesn't consume resources, and the binding $x$ being an unrestricted
$\D$-variable ensures that using it will consume the resources $\D$ the
expression captures, \emph{if} it is used.
% \todo[inline]{Let bindings are hard, if they are used then we use resources. If
% they don't get used then we use no resources! In practice, resources that show
% up in the body of the let must be used, be it by using the let binder, or by
% using them directly. This makes the let binder and the resources in its body
% mutual exclusive.}

@@

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
%
Recursive let bindings also bind expressions lazily and so
introduce a $\D$-bound variable for each binding. Resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-bound variables, or, directly, if the let-bindings are unused.
%
%Additionally, as seen in Section
%\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
%binder in its own definition entail consuming all resources otherwise required
%to type the binder's body.

@@

The typing rule for recursive groups of bindings leverages both the assumption
that all mutually recursive groups are strongly connected and the corollary
that every binder in such a group must be typed with the same linear context,
as observed in Section~\ref{sec:semantic-linearity-examples:recursive-lets}.
Consequently, all bindings of the recursive group are introduced as
$\D$-bound variables
sharing the same usage environment -- using any one of the bindings in a
recursive group entails consuming all resources required to type that same
group -- so, at most a single binder from the group can be used in the
let body.
%\todo{should we say how this means at most once of the binders from the group can be used?} OK
%, which is also why we can use the same linear resources to type each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
because determining a particular $\D$
to type and annotate all binders is not obvious. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

>>>>

% Indeed, usage environments allow us to encode mutual exclusivity between
% alternative ways of consuming linear resources (between $\D$-vars and direct
% resource usage). 
Let-bound variables are the canonical example of a
$\Delta$-bound variable.
% , that is, let-bound variables bind expressions in which
% the resources required to type them are consumed lazily rather than eagerly.
%
Annotating let-bound variables with \ROUNDTWO{their} usage environment $\D$ delays the
consumption of resources to when the variables themselves are used and
guarantees the desired mutual exclusion property between the binder and the resources:
\[
\TypeLet
\]
The rule for (non-recursive) let bindings splits the linear environment in $\D$
and $\D'$. $\D$ is used to type the body $e$ of the let binding $x$. Perhaps
surprisingly, the resources $\D$ are still available to type the let body $e'$, alongside the unrestricted $x$ binding
annotated with the usage environment $\D$. The resources of $e$ being
available in $e'$ reflects how the body of a lazy binder
doesn't consume resources outright, and $x$ being an
\ROUNDTWO{unrestricted $\Delta$-bound variable}
ensures that using $x$ will consume the resources $\D$ that $e$
captures.
% \todo[inline]{Let bindings are hard, if they are used then we use resources. If
% they don't get used then we use no resources! In practice, resources that show
% up in the body of the let must be used, be it by using the let binder, or by
% using them directly. This makes the let binder and the resources in its body
% mutual exclusive.}

@@

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
%
Recursive let bindings also bind expressions lazily and so
introduce a $\D$-bound variable for each binding. \ROUNDTWO{The} resources required to type the
let-bindings are still available in the \ROUNDTWO{let body}. \ROUNDTWO{These
resources may} be consumed
via \ROUNDTWO{the corresponding} $\D$-bound variables, or directly if the let-bindings are unused.
%
%Additionally, as seen in Section
%\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
%binder in its own definition entail consuming all resources otherwise required
%to type the binder's body.

@@

The typing rule for recursive groups of bindings leverages both the assumption
that all mutually recursive groups are strongly connected and the corollary
that every binder in such a group must be typed with the same linear context,
as observed in Section~\ref{sec:semantic-linearity-examples:recursive-lets}.
Consequently, all bindings of the recursive group are introduced as
$\D$-bound variables
sharing the same usage environment -- using any one of the bindings in a
recursive group entails consuming all resources required to type that same
group -- so, at most a single binder from the group can be used in the
let body.
%\todo{should we say how this means at most once of the binders from the group can be used?} OK
%, which is also why we can use the same linear resources to type each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
since \ROUNDTWO{it does not describe how to annotate binders with} a
particular \ROUNDTWO{usage environment} $\D$.
\ROUNDTWO{Our system simply assumes the annotations are explicitly
  provided in the source and checks the correctness of the overall
  recursive let.}
